---
title: 'DTSC 650: Data Analytics In R'
subtitle: 'CodeGrade Final Project Part 2'
output: html_notebook
editor_options:
  chunk_output_type: inline
---

## Student Info

```         
Name: Rhonda Gray
Term: Fall 1
Date:10/13/2024
```

------------------------------------------------------------------------

## General Instructions

------------------------------------------------------------------------

### Academic Integrity

We expect all work conducted at Eastern University to reflect honest and ethical behavior. The primary rule for ensuring academic integrity is that all submitted work must be original and produced by the individual student.

To uphold academic integrity, it is not permitted to view another student's work prior to submitting your own, nor should you provide your work to someone who has not yet submitted their work. Collaborating with other students is only allowed when seeking clarification on a topic or assignment instructions. 

See the syllabus for more information.

### Name of File

Name your assignment file **`Final_Project_Part2.qmd`**. This is a quarto "markdown" file, which has the extension ".qmd".

------------------------------------------------------------------------

### Instructions

For the final section, you will choose four variables to explore. These four variables must be *different* than the three analyzed in Q1-10. Select one of the four variables you chose as the response variable and the other three as predictors. Be sure to read through all of the instructions for Q11-Q15 before choosing your 4 variables.

Feel free to create multiple variables for Q12, Q13, and Q14, e.g. Q12a, Q12b, etc. Please make it clear, though, the distinction between each question/problem with comments, spacing, and new code chunks. If you use Q12a, b, etc., be sure to print the results by doing Q12a or print(Q12a). It is allowable to use `print()` on this part. Your answers must be clearly identifiable.

Take time to tidy your code once you are finished. The easier it is for us to understand, the more credit you could receive.

------------------------------------------------------------------------

### Allowable packages

The only allowable packages are `tidyverse`, `caret`, `Hmisc`, `lsr`, `olsrr`, `psych`, `lm.beta`.

-   If the allowable packages are not installed on your local computer, you'll need to do a one-time installation *from the Console Window in RStudio* for each package like this:

    **`install.packages('<package name>')`**

    *Do not attempt to install packages in code that you submit to CodeGrade.*

-   Note: installing the entire tidyverse with `install.packages('tidyverse')` from the Console Window will save you from having to install any of the tidyverse's individual packages in the future.

-   In your code, load the package's library like this: **`library(<library name>)`**

------------------------------------------------------------------------

### Do / Do not

-   Do use tidyverse functions (e.g., dplyr verbs) for all of the questions where possible.

-   Do use The Pipe, either `%>%` or `|>`.

-   Do use plenty of comments throughout your code chunks so that the grader can follow your line of thinking.

-   Do use the "Q Discussion:" sections provided to explain your thinking and to answer the questions.

-   Do not rearrange data frame outputs unless specified by the question instructions.

-   Do not create multiple copies of the BRFSS dataset in your script. 

-   Do have integrity and submit your own work, not the work of a classmate, an online resource, or someone else.

-   Do not edit the CSV file in Excel, Numbers, Sheets, etc.


------------------------------------------------------------------------

### Submission

You only have ONE ATTEMPT to submit Part 2 to the ACTUAL submission link! On all the other assignments, you got multiple attempts; but not on this assignment.

Ensure you are ready and confident in your work before submitting.

------------------------------------------------------------------------

### Data Set

-   These data come from the [Centers for Disease Control](www.cdc.gov)

-   To answer these questions you will need to use the codebook on Brightspace, called `BRFSS_2021 Codebook`. For part 2 of the project, please note that not all of the variables listed in the codebook are included in the .csv file to be downloaded from Brightspace.

-   Download the `brfss2021.csv` file from Brightspace and place it in the same folder/directory as your script file. Then in RStudio, set your Working Directory to your Source File location: in the menus choose Session \| Set Working Directory \| To Source File Location. You most likely will see some warnings after it loads due to the fact that `read_csv()` will try to guess the column type but because there are so many rows it won't read enough of them to accurately make a guess.

-   You must use the `read_csv()` function when loading the .csv file. Do not use read.csv().

-   Do not rename the .csv file that you download from Brightspace.

-   Do not edit the .csv file.

------------------------------------------------------------------------

### Pipe Notation

You may use the `tidyverse` pipe **`%>%`** or the new base R pipe **`|>`**. For a comparison, see [here](https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/).

You are expected to use pipe notation in all of the CodeGrade assignments. Although there are alternate ways to filter, subset, and summarize data sets, using the pipe creates more readable code and is an important skill to develop.

------------------------------------------------------------------------

### Comments vs. Discussion

You should include both comments within your code chunk as well a discussion after each coded answer. Comments explain your code and your thought process; the discussion uses your code to answer the questions we've asked. We have provided "Q Discussion" sections for each question. Feel free to add more lines in the discussion section.

You can also use markdown headers to organize your discussion and the outline. For more info on markdown, see [here](https://rmarkdown.rstudio.com/authoring_basics.html)

------------------------------------------------------------------------

### Rounding requirement

Unlike Part 1, there is no rounding requirement. Use your best judgment.

------------------------------------------------------------------------

### Dataframe vs. Tibble

Unlike Part 1, there is no dataframe requirement.

------------------------------------------------------------------------

## Questions and Coded Solutions

------------------------------------------------------------------------

### Preliminaries

```{r}
### It's always a good idea when working in RStudio to start with a clean environment. 

### Clear objects from the environment memory that may be leftover from previous versions of your code or other assignments by running the following line:
rm(list = ls())

### Load the libraries you need

library(tidyverse) 
library(psych)
library(lm.beta)





### Load the Data
### Load the data file
### Do not change this line
brf <- read_csv("brfss2021.csv", show_col_types = FALSE)
#remove(brf)

```

------------------------------------------------------------------------

## Questions

------------------------------------------------------------------------

### Q11

Address the values of each of the variables (use the codebook for this and include your answer as comments). For instance, is "none" equal to a value other than 0? Are there extra decimals implied? Are there other values that should be excluded based on the responses they represent? Which variable will you seek to predict from the other variables? Answer these questions in the *Q11 Discussion:* section below the code chunk. Update the dataset to account for these values for the remainder of the questions.

```{r}
### Do not edit the following line. It is used by CodeGrade.
# CG Q11 # 

### TYPE YOUR CODE BELOW ###


brf_part2 <- brf |>
  select(MAXDRNKS, ALCDAY5, BIRTHSEX, INCOME3) |>
as.data.frame()


brf_part2 <- brf_part2 %>%
  filter(
    !BIRTHSEX %in% c(7, 9) &  # Remove unwanted responses 
    !INCOME3 %in% c(77, 99) &  # Remove unwanted responses
    !ALCDAY5 %in% c(777,999,888) &
    !MAXDRNKS %in% c(77,99)
    
  ) %>%
 filter(!is.na(MAXDRNKS) & !is.na(BIRTHSEX) & !is.na(INCOME3))
   
 

#Create variable DKDAY to show number of times they drink per week 

brf_part2 <- brf_part2 %>%
  mutate(
    DKDAY = case_when(
      ALCDAY5 > 0 & ALCDAY5 < 200 ~ round(as.numeric(substr(as.character(ALCDAY5), 2, nchar(as.character(ALCDAY5)))), 2),
      ALCDAY5 >= 200 & ALCDAY5 < 300 ~ round(as.numeric(substr(as.character(ALCDAY5), 2, nchar(as.character(ALCDAY5)))) / 7, 2),
      TRUE ~ ALCDAY5 # Handle other cases
    )
  ) |>
   
  select(DKDAY, everything(), -ALCDAY5)  # Place DKDAY as the first column and remove ALCDAY5

Q11 <- slice_head(brf_part2, n = 10)

Q11 #show first 10 rows of new data

#glimpse(brf_part2)


# unique(brf_part2$DKDAY)


```

### Q11 Discussion:

Your Q11 discussion goes here. Feel free to add more lines as needed.

This project will explore and analyse the effect of income, birth sex  on alcohol consumption levels. It will also look at how max number of drinks on one occasion for an individual effect the weekly alcohol consumption levels. Those who responded "don't know/not sure" or "refused" will be excluded during Q11 step of to streamline the output for analysis. NA's from "MAXDRNKS" and "BIRTHSEX" are removed to clean excessive NA values.  I wanted to use the "ORACE3" as a variable to look at Race but it was not listed in the actual brf dataset so I chose to add "ALCDAY5 to strengthen understanding of drinking patterns.  I created a new variable called "DKDAY" using "ALCDAY5" to show the number of times they drank per week.  This is a measurable number to be used to compare variable relationships.  The 4 variables I will use include "MAXDRNKS","BIRTHSEX" "INCOME3",and  "DKDAY".

"MAXDRNKS" indicates the max number of drinks someone had on any occasion in the last 30 days.
The coding and levels for "MAXDRNKS":

Number
of drinks
77 Don’t
know / Not
sure
99 Refused

The coding and levels for "ALCDAY5" (converted into "DKDAY" *values included below)

1 _ _ Days per
week
2 _ _ Days in
past 30 days
Read if necessary:
A 40-ounce beer
would count as 3
drinks, or a cocktail
drink with 2 shots
would count as 2
drinks.
888 No drinks
in past 30
days
777 Don’t
know / Not sure
999 refused

"DKDAY" shows number of times they drink per week. Value range is 0.14 - 7. This variable is trimmed to only show range 1-7.

"BIRTHSEX" is binary and indicates sex of male or female at the time of birth. 
The coding and levels for "BIRTHSEX":

1 Male
2 Female
7 Don’t
know/Not sure
9 Refused


"INCOME3" notes the annual household income from all sources. 
The coding and levels for "Income3":

01 Less than $10,000?
02 Less than $15,000?
($10,000 to less than
$15,000)
03 Less than $20,000?
($15,000 to less than
$20,000)
04 Less than $25,000
05 Less than $35,000 If
($25,000 to less than
$35,000)
06 Less than $50,000 If
($35,000 to less than
$50,000)
07 Less than $75,000?
($50,000 to less than
$75,000)
08 Less than $100,000?
($75,000 to less than
$100,000)
09 Less than $150,000?
($100,000 to less than
$150,000)?
10 Less than $200,000?
($150,000 to less than
$200,000)
11 $200,000 or more
77 Don’t know / Not sure
99 Refused



### Q12

Remove any outliers for each applicable variable. Make sure you are updating the dataset from Q11 and using this dataset for the remainder of the questions. Briefly explain why you chose the method you used in the *Q12 Discussion* section. Within your code chunk, provide comments explaining what your code is doing.

```{r}
### Do not edit the following line. It is used by CodeGrade.
# CG Q12 # 

### TYPE YOUR CODE BELOW ###

#boxplot(brf_part2$DKDAY)

#unique(brf_partt$DKDAY)

#Create boxplot of MAXDRNKS

#boxplot(brf_part2$MAXDRNKS)

#create boxplot for INCOME3
boxplot(brf_part2$INCOME3)


brf_part2 <- brf_part2 %>% #remove outliers for under 1
  filter(DKDAY >= 1) 

#create upper fence for Income using IQR quantile
upper_fence <- quantile(brf_part2$INCOME3, 0.75) + (1.5 * IQR(brf_part2$INCOME3))

#remove outliers from income
brf_part2 <- brf_part2 |>
  filter (
    INCOME3 < upper_fence
  )
#create upper fence for MAXDRINKS using IQR quantile
upper_fence2 <- quantile(brf_part2$MAXDRNKS, 0.75) + (1.5 * IQR(brf_part2$MAXDRNKS))

#remove outliers from MAXDRNKS
brf_part2 <- brf_part2 |>
  filter (
    MAXDRNKS < upper_fence
  )




 
```

### Q12 Discussion:

Your Q12 discussion goes here. Feel free to add more lines as needed.

I used boxplots to view and analyze the outliers in the data for all four variables. I filtered out the outliers in the INCOME3 and MAXDRNKS variable using the IQR method based on the upper fence because that is where the outlier(s) were located on the boxplot above the 3rd quartile range.   I filtered out the outliers data in DKDAY that was less than one based on my analysis after reviewing the boxplot.  

### Q13

Complete exploratory analyses (for each variable) doing appropriate visualizations with ggplot2. Include visualizations that incorporate a combination of your variables within one plot. Provide an organized discussion of your observations for each of the plots in the *Q13 Discussion* section below.

```{r}
### Do not edit the following line. It is used by CodeGrade.
# CG Q13 # 

### TYPE YOUR CODE BELOW ###

#INCOME3
ggplot(brf_part2, aes(x = "", y = INCOME3)) + 
  geom_boxplot() +
  labs(title = "Boxplot of INCOME3", y = "INCOME3") +
  theme_minimal()

#MaXdRNKs
ggplot(brf_part2, aes(x = "", y = MAXDRNKS)) + 
  geom_boxplot() +
  labs(title = "Boxplot of MAXDRNKS", y = "MAXDRNKS") +
  theme_minimal()

#DKDAY
ggplot(brf_part2, aes(x = "", y = DKDAY)) + 
  geom_boxplot() +
  labs(title = "Boxplot of DKDAY", y = "DKDAY") +
  theme_minimal()


#Birthsex
ggplot(brf_part2, aes(x = "", y = BIRTHSEX)) + 
  geom_boxplot() +
  labs(title = "Boxplot of BIRTHSEX", y = "BIRTHSEX") +
  theme_minimal()


#Boxplot of MAXDRNKS by Income"
ggplot(brf_part2, aes(x = factor(INCOME3), y = MAXDRNKS)) +
  geom_boxplot() +
  labs(title = "Boxplot of MAXDRNKS by Income",
       x = "INCOME3",
       y = "MAXDRNKS") +
  theme_minimal()

# Boxplot of MAXDRNKS by Birth Sex
ggplot(brf_part2, aes(x = factor(BIRTHSEX), y = MAXDRNKS)) +
  geom_boxplot() +
  labs(title = "Boxplot of MAXDRNKS by Birth Sex",
       x = "BIRTHSEX",
       y = "MAXDRNKS") +
  theme_minimal()






```

### Q13 Discussion:

Your Q13 discussion goes here. Feel free to add more lines as needed.

Looking at the "Boxplot of MAXDRNKS by Income", I can see that highest values for max drinks is from middle- class people within the range of income $35,000 - $100,00.  When looking at "Boxplot of MAXDRNKS by Birth Sex" the data shows a relatively slight difference in the median, showing 1 (male) higher. The person who drank the most is a a male at birth. I am surprised at these findings so far because of the ostensibly insignificant of the relationship between birth sex and income. Lowest level of income does show the highest median of max drinks.  It will be interesting to further explore and see the highest predictor variable between sex and income.

### Q14

Run all of the appropriate basic descriptive statistics for each of your variables. What do these statistics reveal? Be sure to address each variable in the "Q14 Discussion" below.

```{r}
### Do not edit the following line. It is used by CodeGrade.
# CG Q14 # 

### TYPE YOUR CODE BELOW ###


#Get Descriptive Stats
summary (brf_part2)

Q14 <- describe(brf_part2)

Q14



```

### Q14 Discussion:

DKDAY: Has a normal distribution with 1.61 sd. the mean and median are close which can indicate even distribution. The kurtosis and skew indicate lighter tails.  This variable has been slightly trimmed (lower fence) and has min of 1 and max of 7, after the removal of outliers.

MAXDRINKS: Median shows that half of the people reported having 3 drinks or less on any occasion in the last 30 days.The mean is greater than the medium which shows some people having considerably more drinks. Max drinks consumed is 13 and min is 1.  mad reveals the data is well spread. Sd is high at 2.66 even though there is a 3.53 trimmed. 

BIRTHSEX: The mean of 1.43 shows slightly higher number of males.  The sd of .49 shows there is a fairly good balance between male and female.

INCOME3: The median being higher than the mean indicates there are more higher income individuals. A slight left- skew, but is still relatively diverse.   

### Q15

Finally, run at least 2 different, appropriate regressions predicting the variable you indicated in Q11. These regressions should use different predictor(s). Organize and explain your code within the code chunk using comments. Identify the best model and provide a discussion of your observations in the *Q15 Discussion* section provided below. Be organized and detailed in your discussion.

```{r}
### Do not edit the following line. It is used by CodeGrade.
# CG Q15 # 

### TYPE YOUR CODE BELOW ###

#convert birthsex as factor
brf_part2$BIRTHSEX <- as.factor(brf_part2$BIRTHSEX)

# Fit the linear regression model for DKDAY using birthsex as predictor
model_BSex <- lm(DKDAY ~ BIRTHSEX, data = brf_part2)

# Fit the linear regression model for DKDAY using income3 as predictor
model_Income <- lm(DKDAY ~ INCOME3, data = brf_part2)


#summarize the models
summary_modSex <- summary(model_BSex)
summary_modIncome <- summary(model_Income)


#Get the AIC for models
aic_modelS <- AIC(model_BSex)
aic_modelI <- AIC(model_Income)

model_SexMax <- lm(DKDAY ~ BIRTHSEX + MAXDRNKS, data = brf_part2)


summary_SexMax <- summary(model_SexMax)
aic_modelSM <- AIC(model_SexMax)


# Predict the likelihood
#predictions <- predict(model_SexMax, newdata = brf_part2, type = "response")

```

### Q15 Discussion:

Which is the strongest predictor for the most alcohol consumption for individuals- sex or income? The results indicate that sex is a stronger predictor than income on alcohol comsumption levels.  2 Ways this determination was made:

1. The stronger linear regression model (model_BSex) using birth sex  as a predictor, shows a lower AIC of 72984.89.
2. The stronger linear regression model(model_BSex) using birth sex as a predictor, shows a higher adjusted r value of 0.004868.

Although birth sex is stronger than income as a predictor, the adjusted r value reveals that it only explains a small fraction of the variance.  Adding a more relevant predictor would likely help strengthen the model fit.

After creating a new model (model_SexMax) to include MAXDRNKS:
Adding in MAXDRNKS to the model lowered the AIC to 72867.99
Adding in MAXDRNKS to the model increased the adjusted r value to 0.01094

These findings show as expected, that the model with 2 predictors is a better fit.  


------------------------------------------------------------------------

# Before submitting to Code Grade:

1)  Clear objects from your environment. Click the broom in the Environment pane in the top right. This will erase any variables (like Q11, Q12) that you've stored.

2)  Rerun all your code. You can click the "Run" option above in this script pane (top right of notebook), then select "Run all". You should have all the variables stored again in the environment pane, and you should see no red error messages in the console below.

3)  **Important**: You only have ONE ATTEMPT to submit Part 2 to the ACTUAL submission link! Ensure you are ready and confident in your work before submitting.

4)  Reread all the directions and questions and make sure you addressed each part carefully, clearly, and thoughtfully.

------------------------------------------------------------------------

This material is for enrolled students' academic use only and protected under U.S. Copyright Laws. This content must not be shared outside the confines of this course, in line with Eastern University's academic integrity policies. Unauthorized reproduction, distribution, or transmission of this material, including but not limited to posting on third-party platforms like GitHub, is strictly prohibited and may lead to disciplinary action. You may not alter or remove any copyright or other notice from copies of any content taken from BrightSpace or Eastern University's website.

© Copyright Notice 2024, Eastern University - All Rights Reserved
